% !TeX root = main.tex
%
% Experimental setup and results for the Pavillon reconstruction.

\section{Experiments}
\label{sec:experiments}

%---------------------------------------------------------------------------
\subsection{Capture and pipeline configuration}
\label{sec:setup}

The subject is a carved wooden ceiling panel, stood vertical and filmed handheld in
a room with an iPhone (3840$\times$2160, 30\,fps, 196\,s). The capture is
single-sided --- the camera sweeps across the panel but never orbits it --- so
parallax between views is small, and the frames alternate between contextual
mid-range shots and extreme close-ups at grazing incidence.

Table~\ref{tab:settings} lists the configuration of the recommended model. Every
value is set in one YAML file; no code changes are required to reproduce it.

\begin{table}[h]
\centering\small
\begin{tabular}{llp{7.2cm}}
\toprule
Stage & Setting & Value / rationale \\
\midrule
Frames & sampling & 4\,fps, capped at 400 frames \\
       & resolution & long edge \textbf{2560}\,px (from 3840; see \S\ref{sec:res}) \\
       & blur filter & variance-of-Laplacian $\geq 50$, measured at a \emph{fixed
                       1600\,px reference scale} so the threshold is
                       resolution-independent \\
       &             & $\rightarrow$ 282 of 400 frames accepted \\
\midrule
SfM    & matcher & exhaustive (low overlap $\Rightarrow$ all pairs) \\
       & mapper  & \textbf{GLOMAP} (global) rather than incremental COLMAP \\
       & result  & \textbf{282/282} registered, 0.985\,px mean reprojection error,
                   152\,792 sparse points \\
\midrule
Split  & strategy & pose-aware; 226 train / 28 val / 28 test \\
\midrule
Train  & backend & \texttt{gsplat}, 30\,k iterations, SH degree 3 \\
       & \textbf{capacity} & \texttt{cap\_max} $=$ \textbf{375\,k} (see \S\ref{sec:capacity}) \\
       & densification & grad threshold $3\times10^{-4}$, interval 100,
                         iterations 500--15\,000, opacity reset every 3000 \\
       & room bounds & AABB(cameras $\cup$ points) $\times1.5$, enforced every 500 steps \\
       & anti-floater & suppress scale $>0.15\cdot$extent; final prune below
                        opacity 0.005 \\
       & depth prior & DepthAnything-v2-Small, Pearson loss, weight 0.1, from iter 2000 \\
\bottomrule
\end{tabular}
\caption{Configuration of the recommended model (\texttt{pavillon\_orbit\_hidetail\_cap375k}).}
\label{tab:settings}
\end{table}

%---------------------------------------------------------------------------
\subsection{Results}

\begin{table}[h]
\centering\small
\begin{tabular}{lrrrrrr}
\toprule
Model & Gaussians & PSNR & median & SSIM & LPIPS & $<$18\,dB \\
\midrule
\multicolumn{7}{l}{\emph{1600\,px dataset, 18 test views}}\\
Baseline (GLOMAP + 3DGS)      & 1\,533\,137 & 23.89 & --- & 0.8249 & 0.2451 & --- \\
A3 regularized                 & 1\,216\,252 & 22.29 & 23.43 & 0.7958 & 0.2901 & 17\,\% \\
A3 ablation (depth off)        & 1\,232\,607 & 22.55 & 22.76 & 0.7978 & 0.2941 & 17\,\% \\
2DGS backend                   &   277\,540 & 13.31 & --- & 0.5778 & 0.8473 & --- \\
\midrule
\multicolumn{7}{l}{\emph{2560\,px dataset, 28 test views}}\\
cap 1.5\,M, ADC                & 1\,244\,835 & 23.08 & 24.48 & 0.8461 & 0.3096 & 11\,\% \\
cap 750\,k, ADC                &   653\,831 & 24.32 & 25.10 & 0.8596 & \textbf{0.3025} & 11\,\% \\
\textbf{cap 375\,k, ADC (recommended)} & \textbf{358\,878} & \textbf{24.92} & 25.13 & \textbf{0.8619} & 0.3132 & \textbf{0\,\%} \\
cap 190\,k, ADC                &   192\,475 & 24.86 & \textbf{25.18} & 0.8559 & 0.3382 & \textbf{0\,\%} \\
$+$ pose refinement            &   354\,250 & 23.94 & 24.40 & 0.8290 & 0.3301 & 0\,\% \\
cap 375\,k, \textbf{MCMC}      &   365\,497 & 24.83 & \textbf{25.32} & \textbf{0.8623} & \textbf{0.3094} & 0\,\% \\
\midrule
\multicolumn{7}{l}{\emph{multi-clip dataset (2 clips), 33 test views}}\\
cap 3.0\,M                     & 2\,502\,172 & 21.36 & 21.23 & 0.8243 & 0.3460 & 33\,\% \\
cap 1.5\,M                     & 1\,225\,089 & 22.03 & 22.82 & 0.8332 & 0.3420 & 24\,\% \\
cap 750\,k                     &   637\,412 & 22.37 & 23.52 & 0.8396 & 0.3350 & 21\,\% \\
\bottomrule
\end{tabular}
\caption{All trained models. \textbf{Rows in different blocks are not directly
comparable}: they are separate datasets with their own SfM solution, held-out split
and image resolution, and PSNR/LPIPS shift with resolution. Comparisons within a
block are valid.}
\label{tab:results}
\end{table}

\paragraph{Resolution.}
\label{sec:res}
Per-view analysis of the 1600\,px model showed the errors were not spread evenly:
every catastrophic view (12.7--18.8\,dB) was an extreme close-up at grazing
incidence, while mid-range views reached 25.6--26.5\,dB. That is a
spatial-frequency deficit --- the pipeline was downscaling a 4K source to 1600\,px,
discarding exactly the carving detail those views demand. Re-running at 2560\,px
with denser frame sampling raised registration to 282/282 (from 181/193) and lifted
median PSNR from 23.43 to 24.48 with SSIM 0.796\,$\rightarrow$\,0.846. The aggregate
mean had concealed this entirely; only the per-view distribution exposed it.

\paragraph{Capacity.}
\label{sec:capacity}
The largest late gain came from making the model \emph{smaller}. Holding everything
else fixed and varying only \texttt{cap\_max}, PSNR and SSIM peak at 375\,k while
the catastrophic-view tail disappears below 750\,k. Cutting the budget $4\times$
from 1.5\,M gained 1.8\,dB and shrank the model $3.5\times$ (295\,MB
$\rightarrow$ 89\,MB). This is sparse-view overfitting: with low parallax the
objective is underdetermined, so surplus primitives buy configurations that fit the
training views and fail on held-out ones (\S\ref{sec:sparse}). The curve genuinely
turns --- 190\,k is worse than 375\,k on PSNR and SSIM --- so 375\,k is an optimum
rather than the smallest budget tried.

The metrics disagree informatively: \textbf{LPIPS is minimised at 750\,k} and
degrades monotonically as capacity shrinks, while PSNR, SSIM and robustness prefer
less. Perceptual detail wants capacity; accuracy and generalisation want less. We
report both operating points rather than collapsing them into one recommendation.

\paragraph{Ablations and negative results.}
\begin{itemize}
  \item \textbf{Depth prior:} disabling it changes test PSNR by only 0.26\,dB
        (22.29\,$\rightarrow$\,22.55). It is nearly free and slightly \emph{increases}
        floater removal, but it is not what determines quality here. The
        $\approx$1.4\,dB cost of the A3 configuration is the final opacity prune, not
        the prior --- an ablation that corrected our initial assumption.
  \item \textbf{Anti-floater:} removes 18--20\,\% of primitives. Baseline models
        carried \textbf{18.2\,\%} of Gaussians below 0.02 opacity with scales up to
        1.0 (scene-spanning); the regularized model has \textbf{0\,\%} sub-0.02 haze,
        max scale 0.16, and median opacity 0.51\,$\rightarrow$\,0.88.
  \item \textbf{2DGS:} collapses to flat renders (13.3\,dB). Disk normals are
        constrained by observing a surface from multiple directions, which a
        single-sided capture does not provide.
  \item \textbf{Multi-clip merging:} SfM merges the clips cleanly (332/332 at
        0.921\,px, 266 train views) and the appearance latents demonstrably absorb
        exposure drift, yet the merged model still trails the best single-clip model
        by $\approx$1\,dB. The second clip enlarges the reconstructed volume while
        adding few views.
  \item \textbf{Pose refinement:} costs 1\,dB. The learned deltas moved only
        $0.036^\circ$ and $7\times10^{-4}$ --- about one pixel --- because GLOMAP's
        poses were already sub-pixel accurate. Refining only training cameras (test
        poses must not be optimised) then misaligns the scene slightly relative to
        the evaluation cameras.
\end{itemize}

\paragraph{Densification strategy.}
Replacing the heuristic adaptive density control with 3DGS-MCMC
\cite{kheradmand2024mcmc} at the same 375\,k budget changes nothing measurable:
the paired per-view difference is $-0.09 \pm 0.45$\,dB (95\,\% CI over 28 held-out
views), i.e.\ indistinguishable. MCMC is ahead on 17 of 28 views and on median,
SSIM and LPIPS; the heuristic wins the mean only through a single $-5.32$\,dB
outlier. We therefore make no quality claim in either direction --- with 28 test
views the experiment cannot resolve differences below roughly half a decibel. The
result is nonetheless informative: it indicates the large gain in
\S\ref{sec:capacity} came from the \emph{budget}, not from how the budget is
allocated, which is what the sparse-view overfitting account predicts. MCMC remains
preferable operationally, since it eliminates the gradient threshold and makes the
budget an explicit target rather than an emergent ceiling.

\paragraph{Two measurement bugs worth reporting.}
Both would have silently corrupted conclusions. First, the blur filter scored
variance-of-Laplacian on the full-resolution frame, a resolution-dependent
statistic: raising the extraction resolution to 2560\,px pushed the median score
from 92 to 36 and rejected 308/400 frames as ``blurry'', which would have made the
higher-resolution run \emph{worse} than the one it replaced. Second, evaluation
rendered without restoring the appearance model, scoring held-out views on the
uncorrected output while training optimised corrected ones; fixing it moved the
multi-clip model from 19.0 to 22.0\,dB. Neither was visible in an aggregate number
--- the first surfaced from a frame-count sanity check, the second from noticing that
PSNR had collapsed while SSIM barely moved, which is the signature of a global
photometric offset rather than broken geometry.

%---------------------------------------------------------------------------
\subsection{Geometry output}

Fusing the composited depth of the recommended model over all 226 training cameras
into a TSDF volume \cite{curless1996tsdf} yields a mesh of 952\,846 vertices and
1\,842\,626 triangles, cropped to the object bounding box. The panel's planks and
carved cross-members are resolved. Two limitations are inherent: volumetric
primitives give a biased expected depth, so the surface is softer than a
surface-aligned method would produce, and the fusion covers whatever the cameras
saw, so the crop is what isolates the subject from the room.
